Towards a standardized evaluation of multiple imputation routines

### 2022-01-25 Manuscript/StandardizedEvaluation.Rmd
Add to do list and some references

@@

Multiple imputation @rubin1987 is a state-of-the-art technique for drawing valid conclusions from incomplete data, and is increasingly used in many substantive fields, such as medicine see e.g. @pulkki2015cumulative, @alkire2015global, @stinesen2015reduced, psychology e.g., @sorbi2014medium, economics e.g., @wong2015economic, @dafoe2013democratic, @allan2015determinants, sociology e.g., @crowley2014flexible, @de2014religiosity and official statistics e.g., @little2012calibrated, @klausch2015selection.

The technique has earned a permanent spot in research and policy making, demonstrated e.g. by the detailed manual created by the National Research Council @little2012prevention. Although a top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers are embracing multiple imputation techniques. After all, the principle of multiple imputation is very intuitive.

@@

To evaluate the ability of an imputation routine to handle the nonresponse model, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand, either by generating data from a statistical model, or by sampling data from a sufficiently large observed set. However, when generating data, two easily overlooked problems may arise.

@@

```{r mis_mech, echo=FALSE, fig.cap="Three missingness mechanisms that yield approximately 50 percent missingness ($N=500$). Displayed are a MCAR mechanism, a right-tailed MAR mechanism generated from bivariate normal data with correlation $\rho=.8$ and the same MAR mechanism, now generated from bivariate normal data with correlation $\rho= 0$."}
knitr::include_graphics("img/plot_mmech.pdf")
```

@@

When MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. When MAR is induced from data without multivariate relations, the inferential implication of the missingness would mimic those of MCAR. This problem is generally overlooked in simulation studies and the procedure to generate missing data is often insufficiently described in resulting publications. Figure \@ref(fig:mis_mech) demonstrates two realizations of missingness mechanisms under MAR mechanisms derived from data with different multivariate relations.

@@

Next, missing data should be induced conform a model that is dependent on the observed data. A straightforward technique for inducing different forms of univariate MAR missingness is described in @fimd, p. 63; see Figure \@(fig:MAR) and a generalization to multivariate MAR missingness can be found in @van2006fully, Appendix B, and @brand1999development, par. 5.2.3. If the missingness is to be induced in longitudinal data, auto-regressive MAR models e.g. cf. @shara2015randomly, model 2 and model 3, can be useful.

```{r MAR, echo=FALSE, fig.cap="Examples of four random missingness mechanisms that yield approximately 50 percent missingness. For negatively correlated missingness covariates the resulting mechanisms are reversed."}
knitr::include_graphics("img/plot_mar.pdf")
```

@@

Fifth, the used missingness mechanism should be detailed, either graphically or written as a function of the data. Often, when inducing missingness, authors remain vague about the actual missingness mechanism under investigation and, even worse, some authors only report something like
\emph{We generated missing data following a MAR missingness mechanism.}
This should be considered unacceptable as claims about the validity of the multiple imputation inference depends heavily on the simulated missingness mechanism.

>>>>

# TO DO

- add ADEMP paper [@morr18], diagnostic checking paper [@zhao22]
- add PPC (refer to Mingyang, and Gelman & King)
- expand convergence section
- add RMSE
- add insights from literature review for chapter on missing data in EHR

@@

Multiple imputation @rubin1987 is a state-of-the-art technique for drawing valid conclusions from incomplete data, and is increasingly used in many substantive fields, such as medicine [see e.g. @pulkki2015cumulative; @alkire2015global; @stinesen2015reduced], psychology [e.g. @sorbi2014medium], economics [e.g. @wong2015economic; @dafoe2013democratic; @allan2015determinants], sociology [e.g. @crowley2014flexible; @de2014religiosity], and official statistics [e.g. @little2012calibrated; @klausch2015selection].

The technique has earned a permanent spot in research and policy making, demonstrated e.g. by the detailed manual created by the National Research Council [@little2012prevention]. Although a top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers are embracing multiple imputation techniques. After all, the principle of multiple imputation is very intuitive.

@@

To evaluate the ability of an imputation routine to handle the non-response model, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand, either by generating data from a statistical model, or by sampling data from a sufficiently large observed set. However, when generating data, two easily overlooked problems may arise.

@@

```{r mis_mech, echo=FALSE, fig.cap="\\label{fig:mis_mech}Three missingness mechanisms that yield approximately 50 percent missingness ($N=500$). Displayed are a MCAR mechanism, a right-tailed MAR mechanism generated from bivariate normal data with correlation $\rho=.8$ and the same MAR mechanism, now generated from bivariate normal data with correlation $\rho= 0$."}
knitr::include_graphics("img/plot_mmech.pdf")
```

@@

When MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. When MAR is induced from data without multivariate relations, the inferential implication of the missingness would mimic those of MCAR. This problem is generally overlooked in simulation studies and the procedure to generate missing data is often insufficiently described in resulting publications. Figure \ref{fig:mis_mech} demonstrates two realizations of missingness mechanisms under MAR mechanisms derived from data with different multivariate relations.

@@

Next, missing data should be induced conform a model that is dependent on the observed data. A straightforward technique for inducing different forms of univariate MAR missingness is described in @fimd, p. 63; see Figure \ref{fig:MAR} and a generalization to multivariate MAR missingness can be found in @van2006fully, Appendix B, and @brand1999development, par. 5.2.3. If the missingness is to be induced in longitudinal data, auto-regressive MAR models e.g. cf. @shara2015randomly, model 2 and model 3, can be useful.

```{r MAR, echo=FALSE, fig.cap="\\label{fig:MAR}Examples of four random missingness mechanisms that yield approximately 50 percent missingness. For negatively correlated missingness covariates the resulting mechanisms are reversed."}
knitr::include_graphics("img/plot_mar.pdf")
```

@@

Fifth, the used missingness mechanism should be detailed, either graphically or written as a function of the data. Often, when inducing missingness, authors remain vague about the actual missingness mechanism under investigation and, even worse, some authors only report something like

> We generated missing data following a MAR missingness mechanism.

This should be considered unacceptable as claims about the validity of the multiple imputation inference depends heavily on the simulated missingness mechanism.
